Spatial variation in predator diel activity patterns

### 2023-07-28 manuscript/manuscript.Rmd
changes following oikos review 1

---
title: Joint spatiotemporal models reveal dramatic shifts in mesopredator diel activity across individuals, habitat types and apex predator activity levels
author:
  - name: Matthew W. Rees
    email: matt.wayne.rees@gmail.com
    affiliation: UOM
    footnote: 1
  - name: Jack H. Pascoe
    email: 
    affiliation: CEC
  - name: Mark Le Pla
    email: 
    affiliation: CEC
  - name: Emma K. Birnbaum
    email: 
    affiliation: CEC
    footnote: 
  - name: Alan Robley
    email: 
    affiliation: ARI
  - name: Nick Golding
    email: 
    affiliation: CU,TKI
  - name: Brendan A. Wintle
    email: 
    affiliation: UOM
  - name: Bronwyn A. Hradsky
    email: 
    affiliation: UOM
    footnote: 
address:
  - code: UOM
    address: Quantitative & Applied Ecology Group, School of Ecosystem and Forest Science, The University of Melbourne, Parkville, VIC, Australia
  - code: CEC 
    address: Conservation Ecology Centre, Otway Lighthouse Rd, Cape Otway, VIC, Australia
  - code: ARI 
    address: Department of Environment, Land, Water and Planning, Arthur Rylah Institute for Environmental Research, Heidelberg, Australia
  - code: CU 
    address: Curtin University, Bentley, WA, Australia
  - code: TKI 
    address: Telethon Kids Institute, Perth Children's Hospital, Nedlands, WA, Australia
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: |
  1. Background.
  2. Methods.
  3. Results.
  4. *Synthesis and applications.*

keywords: "daily activity patterns; camera trap; *Felis catus*; invasive predator; interspecific competition; mesopredator release; diel activity; hierarchical generalised additive models; species interactions; *Vulpes vulpes*."
journal: "Ecography"
date: "`r Sys.Date()`"
bibliography: mybibfile.bib
header-includes:
    - \usepackage{setspace}\doublespacing
    - \usepackage{float}
    - \floatplacement{figure}{H}
linenumbers: true
#numbersections: true
csl: ecography.csl
geometry: margin=1.1in
fontsize: 12pt
output:
  bookdown::pdf_book:
    base_format: rticles::elsevier_article
editor_options: 
  chunk_output_type: console
---
\parskip=12pt

@@

```{r setup, include = FALSE}

knitr::opts_chunk$set(echo = FALSE, cache = FALSE, warning = FALSE, message = FALSE)

INTRODUCTION
==========================

Effective wildlife management relies on understanding species population and behavioural dynamics to accurately predict intervention outcomes. These dynamics depend on both environmental conditions and those of other species, but few statistical approaches incorporate both factors, or allow responses to be dynamic across space and time. This is key because species interactions and niche preferences are multidimensional.

Fear is a near universal driver of animal behaviour. Fear may be from other individuals of their species, a dominant species (including humans) or an event (e.g. fire). Animals may be naive to threats (e.g. alien predators), unable to respond to threats (e.g. scarce prey), or the threat may be too minimal to warrant a response. Some environments more easily facilitate avoidance (e.g. high food and shelter accessibility), and so accounting for landscape context is key. A threat response may be routine, or only employed or enhanced when the threat is heightened. This likely depends on how constant and predictable the threat is.

Animals may respond to threats by altering their activity patterns (i.e. avoidance). An animal may avoid locations where the threat is likely to occur (hereafter ‘spatial avoidance’), avoid times of the day when a threat is likely to occur (hereafter ‘temporal avoidance’), or avoid both locations and times a threat is heightened (hereafter ‘spatiotemporal avoidance’). Such avoidance likely has trade-offs, which may or may not incur costs to individual and population fitness. Modelling approaches which do not allow spatiotemporal and dynamic fear reponses can only make limited inference about species interactions.

Species interactions are usually considered in degrees of overlap between activity patterns. Spatiotemporal avoidance is most often considered, using joint species distribution models (including MSOMs). However, temporal avoidance is increasingly being used (perhaps just allowed due to widely available GPS and camera technology) with kernel density estimation methods. Although, these methods don’t allow environmental covariates, and are particularly susceptible to noisy estimates. Importantly, some are beginning to combine these methods to allow spatiotemporal responses (Azzou et al 2021; Farris et al 2020). However, degrees of overlap in activity patterns only get at routine avoidance, we are generally interested in whether changes in threat levels lead to altered behaviour. This is more indicative of species interactions because static levels of overlap may just reflect different niche preferences rather than avoidance.

Dynamic behavioural responses are also important from a management perspective because they want to know whether an action will cause a change in animal behaviour /  range?). For example, after the reintroduction of wolves in Yellowstone, it was widely thought that this had limited elk populations. Using GPS data and habitat selection methods,  Kohl et al. 2018 found that elk used riskier parts of landscape at times when wolves were least active -- suggesting coexistence without fitness costs. However, there are many other large carnivores in this ecosystem which are likely to impact elk behaviour. And so, will elk respond to wolves in the same way in different environmental contexts? There are little-no examples of replicated spatiotemporal interactions studies.

There is contention around the interactions between two invasive predators - feral cats and foxes. Foxes are routinely culled, and there is concern that this could lead to a mesopredator release of feral cats. Mesopredator release may manifest in both behavioural and numerical changes, which my previous chapter identified - but that behavioral responses (ranging and detectability) were not universal. Here, we aim to dig into the behavioral response further - in particular diel activity times.

In this study, we used generalised additive models to assess (1) how predator (fox and cat) diel activity varies between vegetation types, and (2) whether an introduced mesopredator (cat) changes its diel and/or spatial activity in response to the activity of an introduced apex predator (fox). Because foxes and cats fill a similar function niche, we expected that, on average, they would have similar diel activity patterns across vegetation types (hypothesis 1). However, where fox detections were high, we hypothesised that the subordinate cat to be less active (hypothesis 2) and shift diel activity to times when foxes are least active (hypothesis 3). We tested this on a simple predator system where only cats and foxes occur, and fox activity is manipulated through lethal control. This allowed focus on the sole interactions between these two predators, and ensured differences in correlations wasn’t just due to differences in niche preferences.

@@

MATERIALS AND METHODS
==========================

## Study area

We collated camera-trap data on these two predator species from three distinct studies across two regions of south-west Victoria, Australia (Fig. 1). The purpose of each study was to experimentally survey changes in the mammalian community due to fox control, and so our dataset covered a wide range contexts with and without fox-baiting, allowing us to investigate fox-cat interactions across a gradient of apex predator (fox) activity.

The terrestrial mammalian predator guild is depauperate throughout both regions. Dingoes *Canis familiaris* are now absent throughout, as is the the tiger quoll *Dasyurus maculatus*, more recently absent in the Otway Ranges (last sighted in 2014 despite extensive camera-trapping). Introduced red foxes and feral cats are therefore the only medium-large functional mammalian terrestrial predators across both regions.

In the baited parts of the regions, government land managers conduct ongoing targeted fox control for biodiversity conservation. Poison baits containing 3 mg of sodium mono-fluroacetate (1080) are buried at a depth of 10 cm at 1-km intervals along accessible forest tracks and roads.

In the Glenelg region, Gunditjmara country, natural vegetation is highly fragmented. Here, camera-survey sites were situated in six distinct forest blocks: two are separated by a river, the others are separated by at least *XX* km of pastoral farming and residential property. Foxes in three of the forest blocks have been subject to lethal control since October 2005, with 1080 poison-baits replaced fortnightly [@ROBLEY2014262].

The Otway Ranges, Gadubanud country, is a largely continuous patch of natural vegetation with a strong east-west rainfall and elevation gradient [@swan2021species]. Foxes are controlled through 1080 poison-baits with monthly replacement across most of the Otway Ranges, but a large section to the north-west remains unbaited. The majority of this baiting began in 2016 - 2017, although fox-baiting commenced in some small sections in 2008.

Alongside fragmentation, the extremely rugged, wet forest and rainforest matrix in the western Otway Ranges is the main difference between the two regions. Prescribed fire, primarily for the purpose of asset protection, is another key management action undertaken throughout both regions (except in very wet sections of Otway Ranges).

## Camera-trap surveys

We compiled camera-trap data from three distinct studies across the two regions. After removing camera-traps which were operational for less than 14 days, this totalled 3659 deployments of camera-traps at 1232 camera-trap sites. Survey durations ranged between 14 and 93 days (mean 47), totalling 171,966 trapnights. Camera-trap spacing was variable; the average minimum distance between simultaneously deployed camera-traps was 445, 853 and 1266 metres for each study. Camera-trap were deployed in the Glenelg region between 2013 – 2019, whereas in the Otway Ranges surveys were conducted in 2016 – 2019.

All camera-trap deployments consisted of a Reconyx (Holmen, Wisconsin) brand camera-trap (both white and infrared flashes), attached to a tree or a metal picket, facing a lure. Camera-traps were set-up in two ways depending on the key aim of the project, either targeted toward predators or prey species. One study across both regions positioned camera-traps lower on a tree (around 15 - 30 cm above the ground – angled only slightly downwards) facing a tuna oil lure approximately 2 - 2.5 m away [detailed in @REES2019108287]. The remaining camera-traps were positioned higher on a tree or a metal picket (at least 40 cm above ground) and angled downwards more strongly - facing a lure approximately 1 - 1.5 m away. These lures consisted of peanut butter, golden syrup and rolled oats mixed into a small ball, placed within a tea strainer or PVC pipe container and secured either to the ground, or 20 - 60 cm above ground on a wooden stake. All set-ups were effective in detecting both predator species.

## Data preparation
We first created a table of species detections and deployment information (coordinates, dates) for every camera-trap deployment.  All data analysis was conducted in R version 3.6.3 [@R]. We used lorelograms to identify the minimum interval to approximate independence [@https://doi.org/10.1111/2041-210X.13308]; this indicated 30 minutes discarding repeat detections of a species within 30 minutes was sufficient in reducing temporal autocorrelation. To account for day length variation across space and time, we extracted sunrise and sunset times for each camera-trap deployment using the 'maptools' R-package [@maptools] and adjusted detection times to be relative to sunrise and sunset using the average double anchoring approach described by @https://doi.org/10.1111/2041-210X.13290.

Using the 'reshape' R-package [@reshape], we manipulated the detection table into a dataframe with a row for each hour of the day, for every camera-trap deployment, recording the total number of 'independent' fox and feral cat detections within each hour for the entire camera-trap survey duration.

We identified the broad vegetation type (Ecological Vegetation Class (hereafter 'EVC') group [@delwp2020]) at each camera-trap location. Six vegetation types were surveyed in both regions, although to varying degrees (Table 1). In the Otway region, rainforests are interspersed (primarily in low lying gullies) throughout wet and damp forests at a finescale, and we merged these groups (hereafter 'wet forests').

For each deployment, we calculated the number of poison fox-baits within a 2.3 km radius around each camera-trap - the average maximum distance foxes in these regions travel from their home range centre [@hradsky2017human].

## Generalised additive models 
We modelled predator counts (response variable) with generalised additive mixed-effects models (hereafter 'GAMMs') implemented in the 'mgcv' R-package [@wood2017]. We first fit these models using the poisson family, but switched to the negative binomial family as overdispersion, but not zero-inflation, was detected using the 'DHARMa' R-package [@DHARMa]. We specified a model offset to account for differences in camera-trap survey duration, and a random intercept for each site to account for repeat sampling. We used a duchon spline static spatial smooth to model both spatial autocorrelation and differences in predator activity across space (i.e. the effect of unmodelled environmental covariates). For the fox model, we additionally included a thin plate regression spline for poison-bait density, with a factor smoother for each region, as bait replacement schedules differed between the regions. This formed the base model specification for each model we fitted. Models (n = 2) differed in their specification of the cyclical hour smooth in order to explore (1) variations in predator diel activity across EVC groups, as well as (2) whether feral cat diel activity changed across the gradient of observed fox counts (detailed in the sections below). We used the double penalty model selection approach, which penalises the null space in addition to the range space (i.e. shrinking wiggly terms to linear terms) of the spline basis, meaning covariate effects can be entirely removed from the model [@MARRA20112372]. We plotted models using 'ggplot2' and 'gratia' R-packages [@ggplot2; @gratia].

We expected the activity of each predator species to follow a broadly consistent pattern (e.g. lower during the day), but potentially have small shifts in relative activity between sunset and sunrise across the different EVC groups (1). Fitting separate responses for each EVC group would throw away shared information, which is particularly costly for the least surveyed groups, and would result in more noisy estimates. We therefore specified a global smoother for hour (i.e. average response), along with group-level smoothers with shared wiggliness for each EVC group [i.e. model GS in @pedersen2019hierarchical]. This model structure penalises functions which deviate strongly from the average response, increasing confidence that they are true deviations opposed to the result of sampling noise.

To investigate changes in feral cat  diel activity in response to counts of foxes. We added an extra column to our dataframe with the number of fox detections divided by the log of the number of survey days to account for differences in camera-trap survey durations (hereafter 'adjusted fox counts').

Analysis of stage 1 showed that fox diel activity patterns in wet forests differed relative to all other EVC groups. We therefore expected feral cats might to temporally avoid foxes in a different way in wet forests than the dry vegetation groups, but that responses would be similar in all the drier vegetation groups. We therefore modelled fox-induced changes in feral cat diel activity separately for wet and dry vegetation groups. We further split dry vegetation groups by region, to examine whether patterns remained consistent across space? for spatial replication. We refer to the resulting variable as 'habitat type', which had three levels: wet forests of the Otway Ranges ('wet_otways') dry EVC groups of the Otway Ranges ('dry_otways') and all of the Glenelg region data ('dry_glenelg').

We used a tensor product of hour (using cyclic cubic regression spline) and log adjusted fox count (using thin plate regression spline) smooths to model feral cat diel activity across the range of observed fox activity, separately for each habitat type. By also having double penalty shrinkage turned on, this meant that feral cat activity could vary nonlinearly and be penalised consistently across both dimensions. This model specification allows five different scenarios: that there was (1) no effect of hour or foxes on feral cats (2) a static hourly effect only, (3) a spatial response to foxes only, (4) a spatial response to foxes and an unrelated static hourly effect, (5) a spatial response to foxes and a hourly effect which changes across the range of fox counts.

@@

```{r map, out.width="100%", fig.cap = "Locations of study sites in south-west Victoria, Australia. The grids of camera-traps are denoted by white dots. The Glenelg region is to the west and Otway region to the east. Native vegetation is indicated by dark green, with hill shading. Map tiles by Stamen Design, under CC BY 3.0, map data by OpenStreetMap, under CC BY SA."}
knitr::include_graphics("../figs/fig1_map.png")
```

@@

RESULTS
============

## Predator diel activity across vegetation types

>>>>

---
title: Spatial variation in predator diel activity patterns
author:
  - name: Matthew W. Rees
    email: matt.wayne.rees@gmail.com
    affiliation: UOM,CSIRO
    correspondingauthor: true
  - name: Brendan A. Wintle
    affiliation: UOM
  - name: Jack H. Pascoe
    affiliation: UOM,CEC
  - name: Mark Le Pla
    affiliation: UOM,CEC
  - name: Emma K. Birnbaum
    affiliation: CEC
  - name: Bronwyn A. Hradsky
    affiliation: UOM
address:
  - code: UOM
    address: Quantitative and Applied Ecology Group, School of Ecosystem and Forest Sciences, The University of Melbourne, Parkville, VIC, 3010, Australia
  - code: CSIRO
    address: Health and Biosecurity, Commonwealth Scientific and Industrial Research Organisation, Ecosciences Precinct, Dutton Park, QLD, Australia
  - code: CEC 
    address: Conservation Ecology Centre, Lighthouse Rd, Cape Otway, VIC, 3233, Australia
abstract: |
  *Open research:* Data and code will be deposited on Dryad upon acceptance and can currently be found at this link: https://github.com/matt-w-rees/spatiotemporal-gams-invasive-predators.
keywords: 
  - diel activity patterns 
  - generalised additive model
  - feral cat
  - invasive predator
  - intraguild predator interactions
  - lethal predator control
  - mesopredator release
  - red fox
  - spatiotemporal model
journal: "Oikos; Article"
date: "`r Sys.Date()`"
classoption: preprint, 3p, authoryear
bibliography: "thesis-copy.bib"
header-includes:
  - \usepackage{setspace}\doublespacing
  - \usepackage{lineno}
  - \usepackage{float}
  - \floatplacement{figure}{H}
  - \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}} \setcounter{section}{0} \renewcommand{\thesection}{S\arabic{section}}}
  - \usepackage{pdflscape}
  - \newcommand{\blandscape}{\begin{landscape}}
  - \newcommand{\elandscape}{\end{landscape}}
linenumbers: false
numbersections: true
# Use a CSL with `citation_package = "default"`
csl: "ecography.csl"
fontsize: 12pt
output:
  bookdown::pdf_book:
    base_format: rticles::elsevier_article
    keep_tex: true
    citation_package: default
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo      = FALSE,
  warning   = FALSE,
  message   = FALSE,
  error     = FALSE,
  cache     = FALSE,
  fig.path  = "../figs/",
  fig.align = "center",
  out.width = "80%")

library(kableExtra)
library(dplyr)
library(readr)

```{=tex}
\parskip=12pt 
\newpage
\setcounter{page}{1}
```

# Spatial variation in predator diel activity patterns {.unnumbered}

$~$

\linenumbers

# ABSTRACT {.unnumbered}

Understanding the constraints dominant predators impose on subordinate species is important for predicting ecosystem dynamics and anticipating outcomes of predator management. Subordinate predators may avoid dominant predators in time or space, making it difficult to quantify antipredator behaviours unless joint spatiotemporal analyses are used.
Here, we tested whether an invasive dominant predator (red fox *Vulpes vulpes*) alters the spatiotemporal activity of an invasive subordinate predator (feral cat *Felis catus*). We collated records of both species from 3,667 camera-traps deployed experimentally across two regions of south-eastern Australia with simplified predator guilds; foxes were poison-baited in some landscapes within each region. We used generalised additive models to quantify changes in predator spatiotemporal activity across geographic space, vegetation types, human footprint and (artificially manipulated) gradients of dominant predator activity.
Foxes and cats had similar diel activity patterns when averaged across all sites, however there was important differentiation at a finer scale---cats did not reduce their spatial activity but shifted diel patterns when localised fox activity was high. Cats were crepuscular on average. However, across dry vegetation types of both regions (where foxes were nocturnal), cats shifted to diurnal behaviour with increasing fox activity. In contrast, fox activity was relatively consistent throughout the daily cycle in the wet forest; here cats avoided dawn when fox activity was high. Changes in cat diel activity patterns may facilitate spatial coexistence between these two invasive predators, potentially shifting feral cat impacts onto different native prey.
It is well-appreciated that predator activity varies spatially and fluctuates throughout the daily cycle. However, our study demonstrates that diel activity patterns also vary across space, likely mediated by both landscape-context and fear. dominant predator avoidance appears to be dynamic---a key nuance which is overlooked when simply comparing the average activity overlap between two species.

@@

# INTRODUCTION

Top predators can drive ecosystem structure and dynamics by suppressing populations of rival predators and prey species [@estes2011trophic; @prugh2009rise]. Dominant predators can suppress subordinate species directly through antagonistic interactions and predation, as well as indirectly through fear [@creel2008relationships; @allen2022fear]. Fear-induced behavioural suppression can be as detrimental to subordinate species as predation itself, by limiting resource acquisition and breeding success [@schmitz2004trophic; @preisser2005scared; @brown1999ecology]. Understanding how top predators&mdash;including humans&mdash;constrain the behaviour of subordinate species is therefore important to accurately predict consequences of predator management, such as reintroductions or lethal control [@suraci2016fear; @gaynor2021applied].

Spatial and/or temporal niche partitioning may allow subordinate species to coexist with top predators by reducing encounter-rates and resource overlap [@kronfeld2003partitioning]. However, subordinate species may not consistently employ avoidance behaviours because perceived predation risk is spatiotemporally variable, and antipredator behaviours typically involve a trade-off against resource acquisition, such as limiting or relegating activity to suboptimal places or times [@lima1999temporal; @lima1990behavioral]. Therefore, optimal predator avoidance strategies likely vary across heterogeneous landscapes where resource availability (e.g., shelter, food) and perceived predation risks differ [@kauffman2007landscape; @willems2009predator; @wirsing2021context; @davies2021spatial]. For example, temporal predator avoidance may be preferable over spatial avoidance if food is constantly available throughout the day, and vice versa. These theories are unified under the ecology of fear concept [@brown1999ecology], which has gained increasing attention in recent times [@gaynor2019landscapes]. However, there remains a lack of robust, empirical evidence documenting dynamic changes in antipredator behaviour.

Animal diel activity patterns are increasingly being considered alongside spatial activity and distributional patterns. However, spatial and temporal (in regard to the daily cycle) activity patterns are still often considered in an ad-hoc fashion. For example, by fitting separate models for spatial and temporal overlap, or by repeating spatial analyses (e.g., resource selection functions) at different time periods such night and day [e.g., @basille2015plastic; @kohl2019prey; @smith2019integrating; @wooster2022predator]. Such ad-hoc approaches make it difficult to assess the relative importance, as well as dynamic changes in spatial and temporal processes [@suraci2022beyond].

There are many approaches to modelling intraguild spatiotemporal interactions, such as time-to-encounter models and coefficients of activity overlap [@frey2017investigating]. Time-to-encounter approaches consider fine-scale avoidance or attraction by using time intervals between species detections as the response variable, although do not estimate diel activity patterns [@harmsen2009spatial; @niedballa2019assessing; @fancourt2019introduced]. Multispecies models have been recently developed to assess overlap in spatial and diel activity between two species [@azzou2021sparse], as well as different diel activity patterns based on the presence or absence of another species at a site [@kellner2022two]. Multispecies models are data-hungry and difficult to fit, but can account for imperfect detection and different types of predictor variables [e.g., @parsons2022effect]. Given avoidance behaviours are likely dynamic [@gaynor2019landscapes], there is need for statistical approaches which jointly assess continuous shifts in spatial and diel activity patterns of one species in response to another, particularly when combined with experimental data [@suraci2022beyond].

Generalised Additive Models (hereafter 'GAMs') are increasingly being used to estimate animal diel activity patterns and offer a flexible framework to jointly consider spatial activity. GAMs are useful to model diel activity patterns as they can implement cyclical splines&mdash;joining the end with the start of the day&mdash;as well as complexity penalties to reduce overfitting. If complex nonlinear (i.e., wiggly) effects are not statistically supported, smoothing penalties in GAMs 'shrink' them to a linear (i.e., no wiggliness) effect, or even to a horizontal line - removing the entire effect of the predictor variable and therefore  the need for additional tests of statistical significance (such as p-values and information criteria). GAMS can also capture nonlinear interactions between multiple variables with different units (i.e., 'tensor products'), and offer the ability to share information across categorical variables through hierarchical specifications [@wood2017generalized; @pedersen2019hierarchical]. However, we are only aware of one study which allowed animal diel activity to interact with predation risk as a continuous variable in a GAM [although without considering overall activity, @cunningham2019temporal].

The red fox *Vulpes vulpes* (hereafter 'fox') and feral cat *Felis catus* (hereafter 'cat') have devastating impacts on native prey throughout their introduced range, implicated in the extinction of \~10 and 63 species, respectively [@doherty2016invasive]. The impacts of these invasive predators are particularly extreme on the Australian continent [@woinarski2015ongoing]. Cats are more difficult to manage, and so introduced predator control programs in Australia often target only foxes [particularly through poison-baiting, @reddiex2007control]. As foxes and cats compete for many of the same resources [@fleming2022distinctive], there is concern that lethal fox control could cause mesopredator release [@soule1988reconstructed] of cats [@glen2005complex; @doherty2017stop; @wayne2017recoveries]. However, there is limited evidence that spatial cat activity increases in response to fox control [@hunter2018not], although there are signs of changes in behaviour and population density [@molsher2017mesopredator; @rees2023mesopredator]. Other studies have investigated potential spatial and temporal interactions between these invasive predators [e.g., @roshier2021space], but not in response to fox control, or in a joint spatiotemporal framework that allows flexibility in cat avoidance behaviours with respect to differences in fox activity. Conservation managers are therefore uncertain as to whether fox control may increase the impacts of feral cats on native prey, through increases in cat density or behavioural changes which improve hunting success.

In this study, we first explored (1) how predator diel activity patterns vary across geographic space ('model 1'), and (2) the role of vegetation type and human footprint in predator activity patterns ('model 2'). Using this knowledge to partition our dataset, we then (3) tested whether the activity of a dominant predator (fox) affected the spatiotemporal activity of a subordinate predator (cat; 'model 3'). In line with behavioural avoidance or exclusion, we predicted that cat diel activity pattens would shift towards less-risky times of day (i.e., times when fox activity was lower), and/or spatial cat activity would be lower at sites where fox activity was relatively high. We illustrate how GAMs can provide a simple framework to jointly assess spatial and temporal animal activity patterns, as well as potential species interactions.

@@

# MATERIALS AND METHODS

## Study area and camera-trapping
Our study was conducted in a simple predator system where foxes and cats are the only medium-large (> 1 kg) mammalian predators, and fox activity is manipulated using lethal control in some landscapes. This situation allowed sole focus on the interactions between these two predators, across an experimental gradient of dominant predator (fox) activity. We compiled data from multiple smaller-scale camera-trap studies, each designed to experimentally assess mammal responses to fox control. Overall, we collated 5,449 and 2,202 independent detections of foxes and cats, respectively (separated by at least 30 minutes) from 172,052 camera-trap nights (Table \@ref(tab:diel-tab1)).

We collated camera-trap data across two regions in south-west Victoria, Australia: the Glenelg region and Otway Ranges (Fig. \@ref(fig:diel-map)). Here, dingoes *Canis familiaris* are long-absent throughout, while tiger quolls *Dasyurus maculatus* are long-absent in the Glenelg region and likely functionally extinct in the Otway Ranges (last confirmed sighting in 2014). In broad sections of each region, government land managers conduct ongoing targeted lethal fox control for biodiversity conservation (see sections below for region-specific details). Poison-baits containing 3 mg of sodium fluroacetate (compound 1080) are buried at a depth of 12 - 15 cm at 1-km intervals along accessible forest tracks and roads. Different road densities therefore result in variable densities of poison-baits. Foxes readily dig these poison-baits often resulting in population suppression, whereas it is generally accepted that cats are unlikely to consume buried fox-baits and there is no evidence buried 1080 fox-baits suppresses cat populations [@algar2004feral; @moseby2011use; @risbey1997control; @hunter2018not]. Managers also frequently implement prescribed fire across both regions, primarily to reduce fuel loads to prevent large wildfires.

### Glenelg region
In the Glenelg region, large patches of natural vegetation are fragmented by pastoral farming and residential properties (Fig. \@ref(fig:diel-map)). Foxes in three distinct forest blocks in this region have been subject to poison-baiting since October 2005, with fortnightly bait replacements [@robley2014long]. These forest blocks, along with three similar, unbaited forest blocks to the north have been simultaneously surveyed annually under the 'Glenelg Ark' fox control program since 2005 [40 sites per block, @robley2020glenelg]. Hair-tubes were used to monitor species from 2005 - 2013 [presented in @robley2014long], replaced by camera-traps from 2013; here we present camera-trap data from 2013 - 2019 [@robley2020glenelg]. We also included a further 425 camera-trap deployments at unique locations from early 2018 [M.W.R PhD Surveys', @rees2023data]. This totals 2,039 camera-trap deployments in the Glenelg region, collected in a control-impact experimental design (foxes had been continuously controlled for at 8 - 14 years in the treatment landscapes at the time of these surveys).

### Otway Ranges
The Otway Ranges is a largely continuous patch of natural vegetation with a strong east-west rainfall gradient (Fig. \@ref(fig:diel-map)). A matrix of cool temperate rainforest and wet forest at high altitudes in the south-west descend into a large heathland directly north, and into dry forests and then heathlands to the north-east. Fox-baiting commenced in small sections of the Otway Ranges in 2008 and large-scale systematic baiting began in 2016 - 2017 under the 'Otway Ark' program [@robley2019otway]. For the first six weeks, poison-baits were replaced weekly, then changing to ongoing monthly bait-replacement. There was a pause in baiting for approximately six months during the second half of 2018. Fox control recommenced in late 2018 with four weeks of fortnightly bait-replacement, before returning to monthly bait-replacement. A large section of the Otway Ranges to the north-west remains unbaited, but is monitored as an experimental non-treatment site [@robley2019otway]. Otway Ark managers survey 372 camera-trap sites annually (sequentially across the region); we present one 'before' baiting survey and two 'after' baiting surveys of each site from 2016 - 2018, totalling 1,113 camera-trap deployments [@robley2019otway]. We also include data from an additional before-after control-impact surveys (one 'before' baiting survey and two 'after' baiting surveys) in the western section of the Otway Ranges, conducted annually 2017 - 2019 [@rees2023data]. This added a further 195 sites and 524 camera-trap deployments (Table \@ref(tab:tab-sum)).

### Camera-trap set-ups
All camera-trap deployments consisted of a Reconyx (Holmen, Wisconsin) brand camera-trap (white or infrared flash), attached to a tree or a metal picket, facing a lure. All camera-trap sites were positioned in the forest interior at least 30 m away from roads. The Glenelg Ark and Otway Ark fox monitoring programs positioned camera-traps at least 40 cm above ground on a tree or a metal picket and angled downwards toward a lure approximately 1 - 1.5 m away [@robley2020glenelg; @robley2019otway]. The lures consisted of peanut butter, golden syrup and rolled oats mixed into a small ball, placed within a tea strainer or PVC pipe container and secured either to the ground, or 20 - 60 cm above ground on a wooden stake. The M.W.R PhD surveys across both regions positioned camera-traps lower on a tree (around 15 - 30 cm above the ground) angled only slightly downwards toward a tuna oil lure approximately 2 - 2.5 m away [detailed in @rees2023mesopredator]. Camera-traps were active for an average of 47 days (maximum 93 days), totalling 172,052 trap-nights. Camera-trap spacing averaged 867 m for Glenelg Ark, 1246 m for Otway Ark, and 444 m for the M.W.R PhD surveys (Table \@ref(tab:tab-sum)).

## Data preparation
Analyses were conducted in R version 4.1.3 [@R]. We first used lorelograms to identify the minimum interval to approximate independence [@iannarilli2019lorelograms]; this indicated that discarding repeat detections of a species within 30 minutes was sufficient to reduce temporal autocorrelation. To account for day length variation across space and time, we extracted sunrise and sunset times for each camera-trap deployment using the 'maptools' R-package [@maptools] and adjusted detection times to be relative to sunrise and sunset using the average double anchoring approach described by @vazquez2019comparing. We then built a dataframe consisting of a row for each hour of the day (0 -- 23), for every camera-trap deployment (n = 3,667), recording the total number of 'independent' fox and feral cat detections within each hour across the camera-trap survey.

## Generalised additive models 
We modelled the total number of independent detections of each predator per hour for each camera-trap deployment (response variable) with generalised additive mixed-effect models implemented in the 'mgcv' R-package [@wood2017generalized]. We used the negative binomial family, as overdispersion, but not zero-inflation, was detected with a poisson distribution using the 'DHARMa' R-package [@DHARMa]. We specified the natural log of the number of survey days as a model offset to account for differences in camera-trap survey duration, and a random intercept for each site to account for repeat sampling for models 2 and 3 (as pseudoreplication from repeat sampling in model 1 was explicitly accounted for using the spatial term). For fox models, we also included a smooth effect of poison-bait density [number of baits within a 2.3 km radius around each camera-trap - the average maximum distance adult foxes travel from their home-range centre in these regions, @hradsky2017human] with separate responses per region to account for the effect of fox control. These specifications were included in each model we fitted; models differed in their specification of the cyclical hour smooth to provide inference on variations of predator diel activity across the three questions of interest - this is detailed in the sections below. We used the 'ggplot2' [@ggplot2] and 'gratia' [@gratia] R-packages to plot models.

In this paper, we use the term 'spatial activity' to refer to the number of 'independent' predator detections at a site (offset to account for survey duration; analogous to an activity or abundance index), and 'diel activity pattern' to refer to fluctuations in relative activity throughout the 24-hour daily cycle. The first two models we fit explored how predator spatiotemporal activity varied across these regions, informing the best structure for our third model which tested the activity of foxes limits the spatiotemporal activity of cats (detailed in the sections below).

### How does predator diel activity patterns vary across geographic space? (model 1)
To examine how the spatial activity and diel activity patterns of each predator varied across space, we fit a model for each predator which included a tensor product interaction between a spatial smooth (thin plate regression spline) and hourly smooth. This model specification allowed predators to have different activity levels across space (static across the years surveyed), as well as variation in diel activity pattern across space. Space was modelled using camera-trap coordinates and a duchon spline basis [@miller2014finite].

### How does vegetation type and human footprint affect predator activity patterns? (model 2)
The spatial term in model 1 estimated differences in predator activity across geographic space; we hypothesised that this was partly due to differences in vegetation type, based both on the observed spatial patterns and because vegetation type is a major driver of understorey habitat structure and prey occurrence in these regions [@swan2015predicting; @hradsky2017bayesian]. To test whether the diel activity pattern of each predator varied among vegetation types, we identified the Ecological Vegetation Class group [hereafter 'vegetation type' - standard units for vegetation classification in Victoria, @delwp2020bioregions] for each unique camera-trap site, totalling eight vegetation types. As rainforests are interspersed (primarily in low lying gullies) at fine-scales throughout wet and damp forests in the south-eastern Otway Ranges, we merged them together (hereafter referred to as 'wet forests'). We then estimated predator activity across vegetation types using a hierarchical model specification: a global smoother for hour (i.e., average response) and group-level smoothers with shared wiggliness for the seven vegetation types ['model GS' detailed in @pedersen2019hierarchical]. We also fit a tensor product interaction term for hour and the human footprint index [@venter2016sixteen; @venter2018last]; as human footprint is known to impact animal activity patterns [@gaynor2018influence; @van2023influence]

### Do feral cats avoid foxes in space or time? (model 3)
Fox diel activity across vegetation types showed strong similarity between all vegetation types except wet forests. To examine whether cats avoid foxes in space or time, we therefore modelled fox-induced changes in feral cat diel activity separately for wet forest and dry vegetation types. We further split dry vegetation types by region for further replication. We refer to the resulting variable as 'habitat type', which had three levels: (i) wet forests and rainforests in the western Otway Ranges ('wet_otways'), (ii) dry vegetation types in the Otway Ranges ('dry_otways') and (iii) dry vegetation types in the Glenelg region ('dry_glenelg'). We further hypothesised that cats would avoid foxes in time by becoming more diurnal in dry vegetation types where foxes were mostly nocturnal, but not in wet forests where fox activity showed little variation across the daily cycle.

To investigate changes in feral cat diel activity across the range of observed fox activity, we first quantified fox activity for each camera trap deployment as the total number of fox detections for the deployment divided by the number of survey days, to adjust for differences in survey duration (hereafter 'adjusted fox counts'). We modelled an interaction between hour and adjusted fox counts (as a thin plate regression spline with shrinkage - so complexity penalties could entirely remove this effect from the model if not supported), allowing cats to have nonlinear responses to both hour and adjusted fox counts. We fit separate tensor product interactions for each habitat type (using a 'by-variable' term). For a direct visual comparison to fox activity, we fit another fox model where a diel activity curve was estimated separately across each of the three habitat types.

@@

(ref:fig1cap) Locations of our study regions in south-west Victoria, Australia. The grids of camera-traps are denoted by white dots. The Glenelg region (38°05'54"S 141°44'41"E) is to the west and Otway Ranges (38°57'82"S 141°68'41"E) to the east. Native vegetation is indicated by dark green, with hill shading. *Map tiles by Stamen Design, under CC BY 3.0, map data by OpenStreetMap, under CC BY SA.*.

```{r diel-map, fig.cap = '(ref:fig1cap)'}
knitr::include_graphics("../figs/map_cams.png")
```

@@

# RESULTS
